---
title: 'OHI-Northeast: Plastic Pollution'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/ohi-northeast/src/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---

#Summary

As of now this is copied from Casey's BC trash prep. I've included three different reference points, local max, local 99.99 percentile, and max from the North Atlantic region.

```{r setup,echo=F,message=F,warning=F}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'figs/', message = FALSE, warning = FALSE)

source('~/github/ohi-northeast/src/R/common.R') ### an OHINE specific version of common.R

dir_data_vs <- file.path(dir_M, 'git-annex/bcprep/_raw_data/vanSebille_plastics')
```

#Data

```{r read_in_data}
vs_files_all <- list.files(dir_data_vs, full.names = TRUE) %>%
  setNames(basename(.) %>% str_replace('.csv', ''))

# [1] "gridcell_areas"            "latitudes"                 "lebretonmodel_abundance"  
# [4] "lebretonmodel_mass"        "longitudes"                "maximenkomodel_abundance" 
# [7] "maximenkomodel_mass"       "vansebillemodel_abundance" "vansebillemodel_mass"     

vs_files <- vs_files_all[c(3, 4, 6,7,9,10)] ### select just data layers for models

vs_rasts <- vector('list', length = length(vs_files)) %>%
  setNames(names(vs_files))

for (i in 1:length(vs_files)) { ### i = 1
  vs_file <- vs_files[i]
  vs_fn <- names(vs_files)[i]

  vs_matrix <- read_csv(vs_file, col_names = FALSE) %>%
    data.matrix() %>%
    .[2:181, 1:360]
    ### it appears that row 361 is glitchy; ditch it, and ditch +90 N row as well
  rownames(vs_matrix) <- c(90:-89)
  colnames(vs_matrix) <- c(0:359)

  vs_rast <- raster(vs_matrix)
  extent(vs_rast) <- c(0, 360, -90, 90)
  vs_rast <- rotate(vs_rast)
  crs(vs_rast) <- CRS('+init=epsg:4326')

  vs_rasts[[vs_fn]] <- vs_rast
}

vs_stack <- stack(vs_rasts)
```

##Crop to the Northeast

```{r crop}

vs_stack_crop <- crop(vs_stack, wgs_ext)
plot(vs_stack_crop, col=cols)

### Save raster of van Sebille model by mass
writeRaster(vs_stack_crop[['vansebillemodel_mass']], 
            file.path('int/cw_trash_mass_vansebille.tif'), 
            overwrite = TRUE)

### create North West Atlantic raster: north of Equator; using Law 2010 paper to determine boundaries (Figure 2): http://science.sciencemag.org/content/329/5996/1185.full

n_atl_ext <- extent(c(-90, -45, 5, 50))
vs_mass_natl <- crop(vs_stack[['vansebillemodel_mass']], n_atl_ext)
plot(vs_mass_natl,col=cols)

### Save the global Van Sebille mass map cropped to the North Atlantic for additional reference point options
writeRaster(vs_mass_natl, 
            file.path('int/cw_trash_mass_vansebille_n_atl.tif'), 
            overwrite = TRUE)
```



```{r}

ne_trash <- raster('int/cw_trash_mass_vansebille.tif')%>%
            projectRaster(.,ocean_ne, method = 'ngb')%>%
            resample(ocean_ne,method = 'ngb')%>%
            mask(.,ocean_ne)

plot(ne_trash,col=cols)

writeRaster(ne_trash,filename = 'int/cw_trash_mass_ne.tif',overwrite=T)

```

#Log transform

Not sure we want to do this yet

```{r log_transform}

trash_dens_log <- calc(ne_trash, fun = function(x) {log(x + 1)})

writeRaster(trash_dens_log, file.path('int/cw_trash_mass_log.tif'), overwrite = TRUE)

plot(trash_dens_log, main = 'Mass density\nlog(g/km2)',col=cols)
```

## Rescale marine plastics pressure from zero to 1

Rescale all values to range from zero to 1, using several reference points for comparison.  These rescalings will be done on both log-transformed and raw (linear) scales. The highest amount of trash in the region is located right off New York City (2,144 g/km2). This drives all other values down when a regional reference point is used, whether the data are first log transformed or not.

When using a reference point from the entire North Atlantic region, scores across the Northeast region are, unsurprisingly, lower. The maximum value in the North Atlantic region is off the coast of Venezuela/Colombia (3,881 g/km2).


```{r linear_rescales}

### bring in linear-scaled
trash_dens_lin <- raster(file.path('int/cw_trash_mass_ne.tif'))

### bring in N Atlantic raster
trash_dens_natl <- raster('int/cw_trash_mass_vansebille_n_atl.tif')


trash_ref_9999qtile      <- quantile(trash_dens_lin, prob = 0.9999)
trash_ref_110max         <- max(values(trash_dens_lin), na.rm = TRUE) * 1.1
trash_ref_9999qtile_natl <- quantile(trash_dens_natl, prob = 0.9999)

trash_9999qtile <- calc(trash_dens_lin, fun = function(x) {ifelse(x > trash_ref_9999qtile, 1, x / trash_ref_9999qtile)}) %>%
  setNames('linear_mass')
trash_110max <- calc(trash_dens_lin, fun = function(x) {ifelse(x > trash_ref_110max, 1, x / trash_ref_110max)}) %>%
  setNames('linear_mass')
trash_9999qtile_natl <- calc(trash_dens_lin, fun = function(x) {ifelse(x > trash_ref_9999qtile_natl, 1, x / trash_ref_9999qtile_natl)}) %>%
  setNames('linear_mass')


writeRaster(trash_9999qtile, file.path('int/cw_trash_9999qtile_eez_lin.tif'), overwrite = TRUE)
writeRaster(trash_110max, file.path('int/cw_trash_110max_eez_lin.tif'), overwrite = TRUE)
writeRaster(trash_9999qtile_natl, file.path('int/cw_trash_9999qtile_natl_lin.tif'), overwrite = TRUE)

#plot(trash_9999qtile, main = 'CW trash pressure layer: \n mass density/99.99%ile', col=cols)
plot(trash_9999qtile_natl, main = 'CW trash pressure layer: \n mass density/99.99%ile in N Atlantc', col=cols)

```

```{r log_rescales}

### bring in log-transformed BC map at 1000 m BC Albers projection
trash_dens_log <- raster('int/cw_trash_mass_log.tif')

### bring in N Pacific raster; we are not mapping so scale and projection are unimportant
trash_dens_natl <- raster('int/cw_trash_mass_vansebille_n_atl.tif')
trash_natl_log <- log(values(trash_dens_natl) + 1)


trash_logref_9999qtile      <- quantile(trash_dens_log, prob = 0.9999)
trash_logref_110max         <- max(values(trash_dens_log), na.rm = TRUE) * 1.1
trash_logref_9999qtile_natl <- quantile(trash_natl_log, prob = 0.9999, na.rm = TRUE)

trash_log_9999qtile <- calc(trash_dens_log, fun = function(x) {ifelse(x > trash_logref_9999qtile, 1, x / trash_logref_9999qtile)}) %>%
  setNames('log_mass')
trash_log_110max <- calc(trash_dens_log, fun = function(x) {ifelse(x > trash_logref_110max, 1, x / trash_logref_110max)}) %>%
  setNames('log_mass')
trash_log_9999qtile_natl <- calc(trash_dens_log, fun = function(x) {ifelse(x > trash_logref_9999qtile_natl, 1, x / trash_logref_9999qtile_natl)}) %>%
  setNames('log_mass')


writeRaster(trash_log_9999qtile, filename = 'int/cw_trash_9999qtile_eez_log.tif', overwrite = TRUE)
writeRaster(trash_log_110max, filename = 'int/cw_trash_110max_eez_log.tif', overwrite = TRUE)
writeRaster(trash_log_9999qtile_natl, filename = 'int/cw_trash_9999qtile_natl_log.tif', overwrite = TRUE)

#plot(trash_log_9999qtile, main = 'CW trash pressure layer: \nlog(mass density)/99.99%ile')
plot(trash_log_9999qtile_natl, main = 'CW trash pressure layer: \nlog(mass density)/99.99%ile in N Atlantic')

```
