---
title: 'OHI-Northeast: Plastic Pollution'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/ohi-northeast/src/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---

#Summary

As of now this is copied from Casey's BC trash prep. 

We are going to do a regional bounding box and use the maximum value in that box as the reference point. I'll play with a couple different boxes to look at sensitivity.


```{r setup,echo=F,message=F,warning=F}

knitr::opts_chunk$set(fig.width = 10, fig.height = 10, fig.path = 'figs/', message = FALSE, warning = FALSE)

source('~/github/ohi-northeast/src/R/common.R') ### an OHINE specific version of common.R

dir_data_vs <- file.path(dir_M, 'git-annex/bcprep/_raw_data/vanSebille_plastics')
```

#Data

```{r read_in_data}

vs_files_all <- list.files(dir_data_vs, full.names = TRUE) %>%
  setNames(basename(.) %>% str_replace('.csv', ''))

# [1] "gridcell_areas"            "latitudes"                 "lebretonmodel_abundance"  
# [4] "lebretonmodel_mass"        "longitudes"                "maximenkomodel_abundance" 
# [7] "maximenkomodel_mass"       "vansebillemodel_abundance" "vansebillemodel_mass"     

vs_files <- vs_files_all[c(3, 4, 6,7,9,10)] ### select just data layers for models

vs_rasts <- vector('list', length = length(vs_files)) %>%
  setNames(names(vs_files))

for (i in 1:length(vs_files)) { ### i = 1
  vs_file <- vs_files[i]
  vs_fn <- names(vs_files)[i]

  vs_matrix <- read_csv(vs_file, col_names = FALSE) %>%
    data.matrix() %>%
    .[2:181, 1:360]
    ### it appears that row 361 is glitchy; ditch it, and ditch +90 N row as well
  rownames(vs_matrix) <- c(90:-89)
  colnames(vs_matrix) <- c(0:359)

  vs_rast <- raster(vs_matrix)
  extent(vs_rast) <- c(0, 360, -90, 90)
  vs_rast <- rotate(vs_rast)
  crs(vs_rast) <- CRS('+init=epsg:4326')

  vs_rasts[[vs_fn]] <- vs_rast
}

vs_stack <- stack(vs_rasts)
```

##Compare regional extents

Does the regional extent dramatically change the result of the trash layer? Here are 4 arbitrarily determined extents.

```{r trash_compare_extents}

### create North West Atlantic raster: north of Equator; using Law 2010 paper to determine boundaries (Figure 2): http://science.sciencemag.org/content/329/5996/1185.full

par(mfrow = c(2,2))
library(rworldmap)
world = getMap()

##Extent 1

ext_1 <- extent(c(-82,-56,24,50))
vs_mass_ext1 <- crop(vs_stack[['vansebillemodel_mass']],ext_1)
plot(vs_mass_ext1,col=cols,main = "Extent One",axes=F)
plot(world,col='gray',add=T)

## Extent 2
ext_2 <- extent(c(-90, -45, 5, 50))
vs_mass_ext2 <- crop(vs_stack[['vansebillemodel_mass']], ext_2)
plot(vs_mass_ext2,col=cols, axes=F, main = "Extent 2")
plot(world,col='gray',add=T)

## Extent 3
ext_3 <- extent(c(-81,-45,26,50))
vs_mass_ext3 <- crop(vs_stack[['vansebillemodel_mass']],ext_3)
plot(vs_mass_ext3,col=cols, axes=F, main = "Extent 3")
plot(world,col='gray',add=T)

## Extent 4
ext_4 <- wgs_ext
vs_mass_ext4 <- crop(vs_stack[['vansebillemodel_mass']], wgs_ext)
plot(vs_mass_ext4, col=cols, axes=F, main = "Extent 4")
plot(world,col='gray',add=T)


### Save the global Van Sebille mass map cropped to the North Atlantic for additional reference point options
# writeRaster(vs_mass_natl, 
#             file.path('int/cw_trash_mass_vansebille_n_atl.tif'), 
#             overwrite = TRUE)

invisible(dev.off())
```

##Reference points

Get the maximum value from each extent. Then we can rescale the regional data according to the different values and see the effect.

```{r refs}

ext_list <- list(vs_mass_ext1,vs_mass_ext2,vs_mass_ext3,vs_mass_ext4)

lapply(ext_list,cellStats,stat='max')

```

The maximum across all these extents is either 3,007 which is located off New Jersey, or 3,881 (extent 2 - located off Colombia).

Let's go with 3,007.


## Rescale marine plastics pressure from zero to 1

The highest amount of trash in the region is located right off New York City (3,007 g/km^2^). This drives all other values down when a regional reference point is used, whether the data are first log transformed or not.


```{r trash_rescaled}

trash_ne <- vs_mass_ext4
ref      <- cellStats(trash_ne,stat='max')

#rescale

trash_ne_resc <- trash_ne%>%
                   projectRaster(ocean_ne,method = 'ngb')%>%
                   mask(ocean_ne)%>%
                  calc(.,fun=function(x){ifelse(x>ref,1,x/ref)})

plot(trash_ne_resc,col=cols,main = "Marine Trash",axes=F,
     legend.args=list(text='g/km2', side=3, font=2, line=.5, cex=0.8))

```


***

# Results

## Region scores

By extracting the data for each of the 9 regions using the `zonal` function from the `raster` package we can get the mean score per region.

```{r trash_region_scores}

# extract data for each region:
regions_stats <- zonal(trash_ne_resc,  zones, fun="mean", na.rm=TRUE, progress="text")%>%data.frame()

data <- merge(rgn_data, regions_stats, all.y=TRUE, by.x="rgn_id", by.y="zone") %>%
          dplyr::select(-area_km2)%>%
  gather("year", "pressure_score",starts_with("mean")) 

write.csv(data, "scores/trash.csv", row.names=FALSE)

#map_scores is a function to plot a tmap map of the scores
map_scores(data,'pressure_score',scale_label = "Clean Waters Trash Pressure", map_title = "",rev_col = T)
```





