---
title: 'OHI-Northeast: Economies'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/ohi-northeast/src/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---


``` {r setup, echo=F, message = FALSE, warning = FALSE, results = 'hide'}
knitr::opts_chunk$set(fig.width = 8, fig.height = 6, fig.path = 'figs/', message = FALSE, warning = FALSE)

dir_git <- '~/github/ohi-northeast'
dir_rgn <- file.path(dir_git, 'prep/regions')  ### github: general buffer region shapefiles
source(file.path(dir_git, 'src/R/common.R'))  ### an OHI-Northeast specific version of common.R

library(readxl)
library(DT)
library(dplyr)
library(zoo)

dir_anx <- file.path(dir_M, 'git-annex/neprep')

#round to 2 decimals
options(digits=3, scipen = 999)
```

#Summary

The Economies sub-goal is calculated using coastal GDP data from the [National Ocean Economics Program](http://www.oceaneconomics.org/), and State-wide GDP from the [Bureau of Economic Analysis](http://www.bea.gov/).

***

#Data Sources

##National ocean Economics Program (NOEP)

**Downloaded**: Received from Pat Johston on October 3, 2017.  
**Description**:  GDP [2012 $USD] per sector for RI, ME, MA, CT, NY and NH counties from 2005 to 2014. The data also include number of establishments, jobs and wages for each sector - state - year.  
**Native data resolution**: County level   
**Time range**: 2005 - 2014  
**Format**:  Tabular  

##Bureau of Economic Analysis

**Downloaded**: Accessed October 2, 2017 with R package
**Description**: Regional GDP in chained [2009] $USD for RI, ME, MA, CT and NH  
**Native data resolution**: State level  
**Time range**: 1997 - 2016  
**Format**: Tabular  

https://github.com/us-bea/bea.R

***

# Load data


## Coastal GDP

Two sets of data from NOEP are provided, one for the timeframe 1990 - 2004 and the other from 2005 - 2014. NOEP recommends not using these two together due to inconsistincies across the two, especially in regards to GDP. While both are read in here, we only use the 2005 - 2014 data to calculate the LIV goal for now.

```{r coastal_data}

## Read in raw data which came in .xlsx format with 2 sheets. 

## read in NOEP sheet for all data 1990 - 2004
noep_data_1 <- read_excel(file.path(dir_anx, '_raw_data/NOEP/New_England_ocean_series.xlsx'), sheet = "NOEP")
  
## read in ENOW sheet which contains data for 2005 - 2014 
noep_data_2 <- read_excel(file.path(dir_anx, '_raw_data/NOEP/New_England_ocean_series.xlsx'), sheet = "ENOW")
```

Moving forward just with the more recent data series. Data is provided for each individual coastal county as well as "All Counties" in each state. There is an important distinction between using "All Counties" and simply summing up all state county data. Some data at the county level is supressed due to regulation. But when reported at the State level, this data may be released. We see this when looking at New Hampshires marine transportation data. The sum of NH's two coastal counties for jobs does not equal the reported value for jobs for "All New Hampshire Counties". This is most likely due to data supression at the county level. Therefore we use the sum of all counties rather than the data reported for "All Counties".

We need to read in the MA counties since they are divided across two bioregions
```{r split_MA}

ma_ctys <- read_csv("~/github/ohi-northeast/src/tables/MA_counties.csv")

clean_data <- noep_data_2 %>%
       filter(Sector == 'All Ocean Sectors') %>%
        left_join(rgn_data, by = c("State" = "rgn_name")) %>%
        left_join(ma_ctys, by = c("State", "County")) %>%
        mutate(rgn_id = ifelse(is.na(rgn_id.x), rgn_id.y, rgn_id.x),
               rgn_name = ifelse(is.na(rgn_name), State, rgn_name)) %>%
  select(-rgn_id.x, -rgn_id.y, -area_km2.x, -area_km2.y, -rgn_abrev.x, -rgn_abrev.y) %>%
        mutate(GDP_2012       = ifelse(rgn_name %in% c("Massachusetts-Gulf of Maine", "Massachusetts-Virginian"),
                                       as.numeric(gsub(",","",.$GDP_2012))/2, as.numeric(gsub(",","",.$GDP_2012)))) %>%
  select(State, County, Year, GDP_2012, state, rgn_name, rgn_id)

DT::datatable(clean_data, rownames=F, caption = 'Employment in the Tourism & Recreation sector of Coastal Jobs in New England provided by NOEP')

```



### Meta-analysis

To identify inconsistencies in the data, we take a look at the reported employment values at both the county level and statewide. One would expect that the sum of the county employment values would equal what is reported for the "Statewide" employment values. It seems that this is not the case.

```{r noep_state_vs_county_totals}

states <-  c("Maine", "New Hampshire", "Rhode Island", "Massachusetts", "New York", "Connecticut")

meta <- function(state){
  
  all <- clean_data %>%
    filter(State == !!state,
           str_detect(County, "All")) %>%
    select(Year, GDP_2012) %>%
    rename(all_ctys_gdp = GDP_2012)
  
  out <- clean_data %>%
     filter(State == !!state,
            str_detect(County, "All") == FALSE) %>%
     select(State, County, Year, GDP_2012) %>%
     group_by(Year) %>%
    summarize(totals = sum(GDP_2012, na.rm = T)) %>%
    left_join(all) %>%
    rename(county_totals = totals,
           statewide = all_ctys_gdp) %>%
    gather(key = spatial_res, value = GDP_2012, -Year) %>%
    mutate(State = state)
  
return(out)
}

t <- map_df(states, meta) %>%
  distinct()

ggplot(t, aes(x = Year, y = GDP_2012, color = spatial_res))+
    geom_line() +
    facet_wrap(~State, scales = "free") +
  scale_color_manual(" ", labels = c("Sum of county reported GDP", "Statewide reported GDP"), values = c("blue", "red"))
```

There are some clear discrepancies in the dataset between the total number of jobs reported at the state level ("statewide") and the sum of GDP reported at the county level. Connecticut and Rhode Island numbers almost perfectly align so either datastream would be ok to use. New york and Massachusetts have parallel trends, so using either data stream should result in nearly the same final score as well. New Hampshire has significantly higher GDP beginning in 2011 due to data supression up until that point. We can assume that if data were not suppressed pre-2011 statewide, we would also see parallel trends in New Hampshire as well. Operating under that assumption, we can use the county level data. Maine shows low GDP in earlier years when adding up at the county level. This could be due to a lack of data. For example, Saghadoc county in Maine has no data up until 2010, when the jump happens. This suggests we should use the statewide data for Maine.

**Decision**: Use state level data for Rhode Island, New York, Connecticut, Maine. Use County level data for Massachusetts because we need to split counties across the two MA regions, and use county level data for New hampshire due to the data supression issue.


```{r combine_noep_state_county_data}

#select the data for ME, CT NY, and RI, which is going to use the data reported for "All x counties"
state_data <- clean_data %>%
  filter(str_detect(County, "All"),
         State %in% c("Maine", "Connecticut", "New York", "Rhode Island")) %>%
  select(state = State, year = Year, rgn_id, rgn_name, rgn_gdp = GDP_2012)
  
county_data <- clean_data %>%
  filter(str_detect(County, "All")== FALSE,
         State %in% c("Massachusetts", "New Hampshire")) %>%
  group_by(rgn_id, Year) %>%
  mutate(rgn_gdp = sum(GDP_2012, na.rm = T)) %>% #gdp by region
  select(state = State, year = Year, rgn_id, rgn_name, rgn_gdp) %>%
  distinct()

coast_gdp <- bind_rows(state_data, county_data)

write.csv(coast_gdp, file.path(dir_calc, "layers/eco_coast_gdp.csv"))
```


```{r rgn_coastal_gdp}

ggplot(coast_gdp, aes(x = year, y = rgn_gdp)) +
  geom_line()+
  facet_wrap(~ rgn_name, scales = "free") +
  ylab("2012 $USD") + 
  xlab("Year") +
  ggtitle("Total GDP")

```


***

#Methods

The economies score, $x_{eco}'$, is calculated using the following equation; 

$$x_{eco}' = \frac{\frac{E_{i}}{E_{r}}}{1.025}$$

where $E_{i}$ is the total GDP in year *i*, and $E_{r}$ is the mean annual GDP for the previous year. The growth rate, ${\frac{E_{i}}{E_{r}}}$, is compared to the target GDP growth rate of 2.5% (proportionally, 1.025).


## Calculate annual growth rate

The annual GDP growth rate is calculated by comparing every year's GDP total to the previous year.

```{r gdp_ref_point}

gdp_cst_ref <- coast_gdp %>%
  arrange(year) %>%
  group_by(rgn_id) %>%
  mutate(coast_gdp_1yr = rollapply(rgn_gdp, 1, FUN=mean, align = "right", na.rm=F, partial = T),
         coast_gdp_prev_1yr = lag(coast_gdp_1yr, n = 1)) %>%
  select(state, year, rgn_id, rgn_name, ref = coast_gdp_prev_1yr)

write.csv(gdp_cst_ref, file.path(dir_calc, "layers/eco_coast_gdp_ref.csv"))

```



***


## Setting a GDP target (reference point)

According to the Federal Reserve, a target GDP growth rate is 2-3% per year. We'll set a target of 2.5% each year. We will calculate growth for each year and compare that growth rate to the target of 2.5%. If regions are meeting or exceeding the 2.5% GPD target they will score 100. If not, they will receive a score equal to their growth rate divided by 2.5%.


#Results


```{r eco_status}

  #combining the coastal GDP data and calculate the change
  cst_gdp <- coast_gdp %>%
    left_join(gdp_cst_ref) %>%    #join with the reference point data
    filter(!is.na(ref)) %>%       #remove years with NA for wages (same as removing years pre-2010)
    rename(coast_mean_gdp = ref) %>%
    mutate(cst_chg = rgn_gdp/coast_mean_gdp)

  
  #combine coastal and state data, calculate
  gdp_score <- cst_gdp %>%
    mutate(score = ifelse(cst_chg > 1.025, 1, cst_chg/1.025)*100) %>%
    select(rgn_id, year, score) %>%
    left_join(rgn_data)
  
    ggplot(gdp_score, aes(x = year, y = score, color = rgn_name)) +
    geom_line() +
      ylab("Score")+
      ggtitle("Economies subgoal scores")
```  


***

#References

National Ocean Economics Program. Ocean Economic Data by Sector & Industry., ONLINE. 2012.
Available: http://www.OceanEconomics.org/Market/oceanEcon.asp [3 October 2017]

