---
title: 'Examining SST Rate of Change'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/ohi-northeast/src/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---

Objective: Compare different methods of deriving SST pressure layer. For the Ocean Health Index global, we compare the number of weekly SST anomalies in the most recent 5 years to the # of anomalies in the furthest 5 years (1985-1989). If we looked at rate of change in anomalies and rescaled according to rate, would that be a big difference?

#Setup
```{r setup,echo=F,message=F,warning=F}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'figs/',
                      echo = FALSE, message = FALSE, warning = FALSE)

source('~/github/ohi-northeast/src/R/common.R') ### an OHINE specific version of common.R

library(sp)
library(rgdal)
library(raster)

ssta         = stack(file.path(dir_M,'git-annex/globalprep/prs_sst/data/cortadv5_SSTA.nc'),varname='SSTA')
weekly_sst   = stack(file.path(dir_M,'git-annex/globalprep/prs_sst/data/cortadv5_WeeklySST.nc'),varname='WeeklySST')

names_weekly = names(weekly_sst)
names_ssta   = names(ssta)
```

#Calculate mean annual SST and SST anomalies (SSTA)

For OHI we only calculated standard deviation so we had to calculate mean here.

```{r mean_SST, eval=F}

registerDoParallel(10)

foreach(i = 1982:2012) %dopar%{
  
  s = stack()
  
  for (j in 1:53){
    
    print(j)
    w = which(substr(names_weekly,2,5)==i)[j] 
    if(is.na(w))next()
    
    w_week = weekly_sst[[w]]%>%
      crop(wgs_ext)%>% #crop to NE region
      calc(.,fun=function(x){x * (9/5) - 459.67}) #Kelvin to Farenheit conversion
    
    s = stack(s,w_week)
    
  }
  
  s_mean = calc(s,fun=function(x){mean(x,na.rm=T)},progress='text',filename =paste0('int/sst_annual_means_ne/sst_annual_mean_',i,'.tif'),overwrite=T)
  
}
```

```{r mean_SSTA, eval=F}

#assign 10 cores to parallel processing
registerDoParallel(10)

for(i in 1982:2012){
  print(i)
  s = stack()
  
  for (j in 1:53){
    
    print(j)
    w = which(substr(names_ssta,2,5)==i)[j] 
    if(is.na(w))next()
    
    w_week = ssta[[w]]%>%
      crop(wgs_ext)
    
    s = stack(s,w_week)
    
  }
  
  s_mean = calc(s,fun=function(x){mean(x,na.rm=T)},progress='text',filename =paste0('int/ssta_annual_means_ne/ssta_annual_mean_',i,'.tif'),overwrite=T)
  
}
```

```{r ssta_boxplot}

ssta_annual<-list.files('int/ssta_annual_means_ne',full.names=T)%>%stack()%>%
              calc(.,fun=function(x){x*1.8})%>% #convert change in K to change in F. A 1 K change = 1.8 F change
              projectRaster(ocean_ne)%>%
              mask(ocean_ne)

bwplot(ssta_annual, scales=list(x=list(labels=c(1982:2012), rot=45)), ylab = "Anomaly (degrees F)")

#animate
animate(ssta_annual, zlim = c(-11,30), col=cols)

```

# Original method

Compare the number of weekly anomalies in a 5 year time period as it compares to the historical reference.

```{r}

  l   <- list.files(file.path(dir_M,'git-annex/globalprep/prs_sst/v2015/tmp'),pattern='annual_pos_anomalies',full.names=TRUE)
  
# Grab years 1985-1989 as our reference period
anoms_old <- stack(l[4:8])%>%sum(.)%>%
              crop(wgs_ext)%>%
              projectRaster(ocean_ne)%>%
              mask(ocean_ne)

#grab years 2008-2012 as our current time period
anoms_new <- stack(l[27:31])%>%sum(.)%>%
              crop(wgs_ext)%>%
              projectRaster(ocean_ne)%>%
              mask(ocean_ne)

anoms_diff <- raster('int/sst_anom_diffs/sst_anom_diffs_2008-2012.tif')

#plot the three
par(mfrow = c(3,2))
plot(anoms_old,axes=F,col=cols,main = 'Number of weeks with SST anomalies \n 1985-1989')
hist(anoms_old)
plot(anoms_new,axes=F,col=cols,main = 'Number of weeks with SST anomalies \n 2008-2012')
hist(anoms_new)
plot(anoms_diff,axes=F,col=cols,main = 'Difference in number of anomalies')
hist(anoms_diff)
invisible(dev.off())

#The rescaled layer using this method using hte maximum change in anoms globally as ref point
anom_resc <- raster('int/sst_rescale/sst_rescale_2008-2012.tif')
par(mfrow=c(1,2))

plot(anom_resc,col=cols,axes=F)
plot(rgns,add=T,main = 'Rescaled Pressure Layer using \n Difference in Anomalies Method')
hist(anom_resc)

```

# SST Rate of change

Calculates the trend per pixel from 1982 - 2012 for SST

```{r sst_rate_allyrs}

#create a raster stack of all annual SST mean rasters
annual_stack <- list.files('int/sst_annual_means_ne',full.names=T)%>%stack()

#get a single raster output with rate of change
time <- c(1:31) #31 years
fun  <- function(x) { if (is.na(x[1])){ NA } else {lm(x ~ time)$coefficients[2] }} 

rate <- calc(annual_stack, fun, progress = 'text')%>%
          projectRaster(crs = us_alb)%>%
          resample(ocean_ne,method = 'ngb')%>%
          mask(ocean_ne)

plot(rate,axes=F,col=cols,main = "Rate of Change SST \n deg. F 1982-2012")
hist(rate)
```


# SST Difference in rate of change

Look at rate of change in only the last 10 years (2003-2012) compared to rate of change 1982-1991


```{r sst_rate_diff_10yrs}

#rec = recent
rec_stack <- annual_stack[[22:31]]
old_stack <- annual_stack[[1:10]]

#get a single raster output with rate of change
time <- c(1:10) #10 years

rec_rate <- calc(rec_stack, fun, progress = 'text')
old_rate <- calc(old_stack, fun, progress = 'text')

invisible(dev.off())

par(mfrow = c(2,2))
plot(rec_rate,col=cols,zlim = c(-2.1,2.6), axes=F, main = "Rate of Change SST \n deg.F 2003-2012")
hist(rec_rate)
plot(old_rate,col=cols,zlim = c(-2.1,2.6), axes=F, main = "Rate of Change SST \n deg.F 1982-1991")
hist(old_rate)

diff <- rec_rate - old_rate
invisible(dev.off())
par(mfrow=c(1,2))
plot(diff,col=cols)
hist(diff)
m <- cellStats(diff,stat='mean') 

```

This plot shows that over the years 2003-2012, the NE region is warming at a faster rate, `r m` degrees faster per year, than from 1982-1991. 

To see how different the outcome is if we compare the first and last 5 years, we do the same thing but using 1982-1986 and 2008-2012. 

```{r sst_rate_diff_5yrs}

#first 5 years compared to recent 5 years
rec_stack <- annual_stack[[27:31]]
old_stack <- annual_stack[[1:5]]

#get a single raster output with rate of change
time <- c(1:5) #5 years

rec_rate <- calc(rec_stack, fun, progress = 'text')

old_rate <- calc(old_stack, fun, progress = 'text')

invisible(dev.off())

par(mfrow = c(2,2))
plot(rec_rate,col=cols,zlim = c(-6.3,6.1), axes=F, main = "Rate of Change SST \n deg.F 2008-2012")
hist(rec_rate)
plot(old_rate,col=cols,zlim = c(-6.3,6.1), axes=F, main = "Rate of Change SST \n deg.F 1982-1987")
hist(old_rate)

diff <- rec_rate - old_rate
invisible(dev.off())
par(mfrow=c(1,2))
plot(diff,col=cols)
hist(diff)
m <- cellStats(diff,stat='mean') 

```

This plot shows that over the years 2008-2012, the NE region is warming at a faster rate, `r m` degrees faster per year, than from 1982-1986. 

# SST Rescale using difference in rate - 5 years
```{r rescale_sst_5yr_diff}

ref <- cellStats(diff,stat='max')

diffs_resc <- calc(diff,fun=function(x){ifelse(x<0,0,x/ref)})
plot(diffs_resc,axes=F,col=cols)
hist(diffs_resc)
invisible(dev.off())
```

***

I ran the same exploration but with SSTA (sea surface temperature anomalies)

# SSTA Rate of change

Calculates the trend per pixel from 1982 - 2012

```{r ssta_rate_allyrs}

#get all annual mean SSTAs
ssta_stack <- list.files('int/ssta_annual_means_ne',full.names=T)%>%
                  stack()%>%
                  projectRaster(ocean_ne)%>%
                  mask(ocean_ne)

#get a single raster output with rate of change
time <- c(1:31) #31 years

#calculate rate of change over time
ssta_rate <- calc(ssta_stack, fun, progress = 'text')

plot(ssta_rate,axes=F,col=cols,main = "Rate of Change SST Anomalies \n deg. F 1982-2012")
hist(ssta_rate)
m <- cellStats(ssta_rate,stat='mean') #0.033 degrees per year
```

The mean change in SST anomaly is `r m` degrees per year

***
# SSTA Difference in rate of change

Look at rate of change in only the last 10 years (2003-2012) compared to rate of change 1982-1991?

```{r ssta_rate_diff_10yrs}

#rec = recent
rec_stack <- ssta_stack[[22:31]]
old_stack <- ssta_stack[[1:10]]

#get a single raster output with rate of change
time <- c(1:10) #10 years

rec_rate <- calc(rec_stack, fun, progress = 'text')

old_rate <- calc(old_stack, fun, progress = 'text')

invisible(dev.off())

par(mfrow = c(2,2))
plot(rec_rate,col=cols,zlim = c(-2.1,2.6), axes=F, main = "Rate of Change SSTA \n deg.F 2003-2012")
hist(rec_rate)
plot(old_rate,col=cols,zlim = c(-2.1,2.6), axes=F, main = "Rate of Change SSTA \n deg.F 1982-1991")
hist(old_rate)

diff <- rec_rate - old_rate
invisible(dev.off())
par(mfrow=c(1,2))
plot(diff,col=cols,axes=F,main = "Difference in rate of SSTA \n 1982-1991 vs 2003-2012")
hist(diff)
invisible(dev.off())
m <- cellStats(diff,stat='mean') #0.18

```
This plot shows that over the years 2003-2012, the NE region is warming at a faster rate, `r m` degrees faster per year, than from 1982-1991. 

To see how different the outcome is if we compare the first and last 5 years, we do the same thing but using 1982-1986 and 2008-2012. 

```{r ssta_rate_diff_5yrs}

#first 5 years compared to recent 5 years
rec_stack <- ssta_stack[[27:31]]
old_stack <- ssta_stack[[1:5]]

#get a single raster output with rate of change
time <- c(1:5) #5 years

rec_rate <- calc(rec_stack, fun, progress = 'text')

old_rate <- calc(old_stack, fun, progress = 'text')

invisible(dev.off())

par(mfrow = c(2,2))
plot(rec_rate,col=cols,zlim = c(-6.3,6.1), axes=F, main = "Rate of Change SST \n deg.F 2008-2012")
hist(rec_rate)
plot(old_rate,col=cols,zlim = c(-6.3,6.1), axes=F, main = "Rate of Change SST \n deg.F 1982-1987")
hist(old_rate)

diff <- rec_rate - old_rate
invisible(dev.off())
par(mfrow=c(1,2))
plot(diff,col=cols,axes=F,main = "Difference in rate of SSTA \n 1982-1987 vs 2008-2012")
hist(diff)
m <- cellStats(diff,stat='mean')

```

This plot shows that over the years 2008-2012, the NE region is warming at a faster rate, `r m` degrees faster per year, than from 1982-1986. 

# SSTA Rescale using difference in rate - 5 years

```{r rescale_ssta_5yr_diff}
ref <- cellStats(diff,stat='max')

diffs_resc <- calc(diff,fun=function(x){ifelse(x<0,0,x/ref)})
plot(diffs_resc,axes=F,col=cols)
hist(diffs_resc)
invisible(dev.off())
```


