---
title: 'OHI-Northeast: Sea Surface Temperature Pressure'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/ohi-northeast/src/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---

``` {r child = file.path(dir_git, 'src/templates/ohibc_prov_ftr.Rmd')}
```

``` {r setup, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'figs/',
                      echo = FALSE, message = FALSE, warning = FALSE)

source('~/github/ohi-northeast/src/R/common.R') ### an OHINE specific version of common.R

library(sp)
library(rgdal)
library(raster)


### goal specific folders and info
goal      <- 'tr'
scenario  <- 'v2016'
dir_goal  <- file.path(dir_git, 'prep', goal, scenario)
dir_goal_anx <- file.path(dir_anx, goal, scenario)

```

#Summary
[general description: What data are being generated? Why (what project, etc.)? Upstream/downstream processing information that might be helpful?  Other information?]

***

#Data Source [NOTE: can be copied from README.md in rawdata file]
**Reference**: [citation for source data; website, literature, contact information. Version of data (if relevant). Screenshots if a series of menus had to be navigated to obtain the data.]

**Downloaded**: [date downloaded or received]

**Description**:  [e.g., surface aragonite state]

**Native data resolution**: [e.g., 1 degree, 30 m, etc.]   

**Time range**: [e.g., 1880-1899, monthly data provided for each year] 

**Format**:  [e.g. NetCDF]

***
  
#Methods

Read in the global data used for OHI 2016.
```{r global_data}

#each of these rasters is the difference in SST anomalies between a given 5 year period and the historical reference period (1985-1990)
sst_global_files <- list.files('~/github/ohiprep/globalprep/prs_sst/v2016/int', pattern = '.tif',full.names=T)

```

Crop all files to the northeast region

```{r crop}

#get difference in anomalies for the NE region

# lapply(sst_global_files, cropPressure,rgn_crs = us_alb, rgn_ext = ne_ext, filename = paste0('int/sst_anom_diffs_',substr(sst_global_files,77,85),'.tif'))

registerDoParallel(8)

foreach(f = sst_global_files) %dopar%{

  raster(f)%>%           #raster the file
    crop(wgs_ext)%>%     #crop to the WGS extent 
    projectRaster(crs = us_alb)%>% #projectRaster to US albers (crop first to reduce time in reprojecting)
    crop(ne_ext, filename = 
           paste0('int/sst_anom_diffs_',substr(f,77,85),'.tif'),overwrite=T) #crop the projected raster to our NE extent
    
}

plot(raster('int/sst_anom_diffs_2008-2012.tif'),axes=F,col=cols,box=F)
plot(ne_rgns,add=T)
```

Visualize change over time

```{r}

Reference point

Mask and rescale





***

#Citation information  
[citation information: include if these data will have their own specific citation.]