---
title: 'OHI-Northeast: Biodiversity: Species sub-goal'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/ohi-northeast/src/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---

``` {r setup, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'Figs/',
                      echo = FALSE, message = FALSE, warning = FALSE)

library(sp)
library(rgdal)
library(raster)
library(DT)

source('~/github/ohi-northeast/src/R/common.R')  ### an OHINE specific version of common.R

dir_git <- '~/github/ohi-northeast'
#dir_rgn <- file.path(dir_git, 'prep/regions')  ### github: general buffer region shapefiles
dir_anx <- file.path(dir_M, 'git-annex/neprep')


### goal specific folders and info
scenario    <- 'v2017'
goal        <- 'bio'
dir_git     <- '~/github/ohi-northeast'
dir_goal    <- file.path(dir_git, 'prep', goal)
dir_spatial <- file.path(dir_git, 'prep/_spatial')

dir_goal_anx        <- file.path(dir_M, 'git-annex/neprep/bio', scenario) 
dir_goal_global     <- file.path('~/github/ohiprep/globalprep/spp_ico', scenario)
dir_goal_anx_global <- file.path(dir_M, 'git-annex/globalprep/spp_ico', scenario)

## Source goal specific functions

source('functions.R')

```


# Summary: OHI-Northeast Species Subgoal (Biodiversity)

This script prepares scores (status and trend) for species richness in 
the US Northeast's coastal regions.  Spatial data from IUCN and Aquamaps is
combined with extinction risk information from IUCN.

Currently, the Species Richness sub-goal model is identical to the OHI Global 
model: a region's status is based upon an area-weighted average of species
health across each reporting region.

From Halpern et al (2012):

> The target for the Species sub-goal is to have all species at a risk status of Least Concern. We scaled the lower end of the biodiversity goal to be 0 when 75% species are extinct, a level comparable to the five documented mass extinctions and would constitute a catastrophic loss of biodiversity. The Status of assessed species was calculated as the area- and threat status-weighted average of the number of threatened species within each 0.5 degree grid cell.

For the Northeast assessment, our calculation will be slightly different, though the end result will be identical.

**Species area-weighted risk:**  For each species within a region, the risk score is weighted by the proportion of the species' range within the given region.  To determine the mean area-weighted risk, the area-weighted risks are summed and divided by the total number of species within the region.

$$R_{spp/rgn} = (Risk)*\frac{\displaystyle\sum_{rgn}(A_{cell} * pA_{cell/rgn})}{A_{rgn}}$$

$$\bar{R}_{spp} = \frac{\displaystyle\sum_{species}(R_{spp/rgn})}{n_{spp}}$$

**Species goal model**

$$X_{SPP} = \frac{((1 - \bar{R}_{spp}) - 0.25)}{(1 - 0.25)} * 100%$$

where:

* $X_{SPP}$ is Species goal status
* $R_{spp/rgn}$ is area-weighted extinction risk for one species within a region
* $\bar{R}_{spp}$ is area-weighted mean extinction risk for a region
* $A_{cell}$ is cell area
* $pA_{cell-rgn}$ is percent of cell area included in region
* *Risk* is scaled value for species extinction risk category, based on: 
    * 'LC' = 0.0, 'NT' = 0.2, 'VU' = 0.4, 'EN' = 0.6, 'CR' = 0.8, 'EX' = 1.0
* SPP trend is calculated by examining the linear trend of mean extinction risk category, based upon the time series of risk categories from the IUCN Red List.  This calculation is performed in functions.R.

-----

# Data sources

AquaMaps

* From http://www.aquamaps.org/main/home.php: "AquaMaps are computer-generated predictions of natural occurrence of marine species, based on the environmental tolerance of a given species with respect to depth, salinity, temperature, primary productivity, and its association with sea ice or coastal areas."
* Citation: Kaschner, K., J. Rius-Barile, K. Kesner-Reyes, C.Garilao, S.O. Kullander, T. Rees and R. Froese (2015). AquaMaps: Predicted range maps for aquatic species. World wide web electronic publication, www.aquamaps.org, Version 08/2015.

IUCN Red List spatial data: species range map shapefiles

* Species range map shapefiles downloaded from: http://www.iucnredlist.org/technical-documents/spatial-data
* Citation: IUCN 2016. The IUCN Red List of Threatened Species. Version 2015.3. <http://www.iucnredlist.org>. Downloaded on 9 December 2016.

IUCN Red List species index: list of all IUCN red list species, incl IUCN species ID and extinction risk category

* Extinction risk categories, past assessments, and habitat information, by species, downloaded from: http://apiv3.iucnredlist.org/
* Citation: IUCN 2016. The IUCN Red List of Threatened Species. Version 2016-3. <http://www.iucnredlist.org>. Downloaded <date of last script process>.

-----
  
#Methods

1. Get all AquaMaps species in the region by getting cell ids for the region then filtering aquamaps dataset on those cells. Keep the species name and info though
2. Do the same ^ but with IUCN data 
3. Replace those species in this NE AquaMaps collection with maps from the data portal


## Define assessment regions 

Using OHI-Northeast region polygons, determine 0.5° raster cells corresponding to OHI-Northeast and to each region.  Save raster to local directory for later use.

``` {r setup_region-to-cell_lookup}

rgn2cell_file     <- file.path(dir_goal, 'data/rgn2cell.csv')
loiczid_rast_file <- file.path(dir_goal, 'data/loiczid.tif')

if(!file.exists(rgn2cell_file) | !file.exists(loiczid_rast_file)) {
  
  ## Read in region polygons, transform to WGS84 CRS
  poly_ne_rgn <- rgns%>%
    spTransform(p4s_wgs84)
  
  rgn2cell_list <- spp_rgn2cell(poly_ne_rgn) ###STUCK HERE
  
  write_csv(rgn2cell_list[[1]], rgn2cell_file)
  writeRaster(rgn2cell_list[[2]], loiczid_rast_file, overwrite = TRUE)
}
# } else {
#   
#   git_prov(c(rgn2cell_file, loiczid_rast_file), filetype = 'output')
#   
# }

```


## Identify NE-specific species

* Data-set specific idiosyncracies:
    * For AquaMaps, we select a threshold (currently 0%) to set the minimum probability of occurrence that determines species "presence."
    * For IUCN, no threshold is needed; but the shapefiles include a "presence" attribute in which a value of 5 indicates a region in which a subpopulation has become extinct.  We use this to manually reset local extinction risk to EX.
    * Note that for IUCN, we determine the proportional area when extracting polygons; currently we just consider any presence to fill the cell (similar to assuming even a low AquaMaps probability to indicate presence within the entire cell) - so proportional area is ignored in the calculations.
* Data outputs: 
    1. Species (`am_sid` or `iucn_sid`) presence by cell id (`loiczid`), including `cell_area`, `rgn_id`, and proportion of cell within the region (`prop_area`).  For IUCN, also includes `subpop`.
        * `am_spp_area_by_rgn.csv` and `iucn_spp_area_by_rgn.csv`
    2. Species (`am_sid` or `iucn_sid`) by region (`rgn_id`), including total area in each OHIBC region (`spp_area`) and proportional area within each region (`spp_pct_area`). For IUCN, also includes `subpop`.
        * `am_spp_cells_bc.csv` and `iucn_spp_cells_bc.csv`

``` {r identify_am_ohibc_species}

reload <- FALSE

### collect AquaMaps species local to OHI NE

am_spp_cells_file <- file.path(dir_goal, 'int/am_spp_cells_bc.csv')
am_spp_rgn_file   <- file.path(dir_goal, 'int/am_spp_area_by_rgn.csv')

if(!file.exists(am_spp_rgn_file) | reload) {
  
  rgn2cell_df <- read_csv(file.path(dir_goal, 'spatial/rgn2cell.csv')) %>%
  group_by(rgn_id) %>%
  mutate(area_tot = sum(cell_area * prop_area)) %>%
  ungroup()

  ### Load Aquamaps species per cell table
  am_spp_cells_global_file <- file.path(dir_goal_anx_global, 'int/am_cells_spp_prob0.csv')

  am_spp_cells <- read_csv(spp_cell_file) %>%
    select(am_sid = speciesid, loiczid) ### drop probability column - use all cells!

  ### filter out to just cells in BC regions, then summarize area by species ID and region
  message('Trimming AquaMaps global species list to local species only, then determining area per region...')
  am_spp_cells_bc <- rgn2cell_df %>%
    left_join(am_spp_cells, by = 'loiczid')
  
  am_spp_rgn_area <- am_spp_cells_bc %>%
    group_by(am_sid, rgn_id) %>%
    summarize(spp_area     = sum(cell_area * prop_area),
              spp_pct_area = sum(cell_area * prop_area / area_tot))

  message(sprintf('Writing Aquamaps species area per region file to: \n  %s', am_spp_rgn_file))
  write_csv(am_spp_cells_bc %>% select(-area_tot, -rgn_name), am_spp_cells_file)
  write_csv(am_spp_rgn_area, am_spp_rgn_file)
  
  # head(am_spp_rgn_area)
  # head(am_spp_cells_bc %>% select(-area_tot, -rgn_name))
  
} else {
  
  git_prov(am_spp_rgn_file, filetype = 'output')
  
}
```

``` {r get_iucn_ohibc_spp}

### collect IUCN species local to BC

iucn_spp_cells_file <- file.path(dir_goal, 'int/iucn_spp_cells_bc.csv')
iucn_spp_rgn_file <- file.path(dir_goal, 'int/iucn_spp_area_by_rgn.csv')

if(!file.exists(iucn_spp_rgn_file) | reload) {
  
  rgn2cell_df <- read_csv(file.path(dir_goal, 'spatial/rgn2cell.csv')) %>%
  group_by(rgn_id) %>%
  mutate(area_tot = sum(cell_area * prop_area)) %>%
  ungroup()

  iucn_spp_cells_global_file <- file.path(dir_goal_anx_global, 'int/iucn_cells_spp.csv')

  iucn_spp_cells <- read_csv(iucn_spp_cells_global_file)

  iucn_spp_cells_bc <- rgn2cell_df %>%
    rename(prop_area_rgn = prop_area) %>%
    left_join(iucn_spp_cells %>%
                select(-prop_area),
              by = 'loiczid') %>%
    filter(presence != 5) %>%
    select(-presence) %>%
    distinct()
  
  iucn_spp_rgn_area <- iucn_spp_cells_bc %>%
    group_by(iucn_sid, rgn_id, subpop) %>%
    summarize(spp_area     = sum(cell_area * prop_area_rgn),
              spp_pct_area = sum(cell_area * prop_area_rgn / area_tot)) %>%
    ungroup()
  
  message('Writing IUCN species area per region file to ', iucn_spp_rgn_file)
  write_csv(iucn_spp_cells_bc %>% select(-area_tot, -rgn_name), iucn_spp_cells_file)
  write_csv(iucn_spp_rgn_area, iucn_spp_rgn_file)
  
  # head(iucn_spp_rgn_area)
  # head(iucn_spp_cells_bc %>% select(-area_tot, -rgn_name))

} else {
  
  git_prov(iucn_spp_rgn_file, filetype = 'output')
  
}

```









***

#Citation information  
[citation information: include if these data will have their own specific citation.]