---
title: 'OHI-Northeast: [NAME OF GOAL/LAYER]'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/ohi-northeast/src/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---



include the above as the start of a .Rmd; include the below as the last lines.

``` {r child = file.path(dir_git, 'src/templates/ohibc_prov_ftr.Rmd')}
```

here is a setup code chunk with many of the common parameters:
``` {r setup, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'Figs/',
                      echo = FALSE, message = FALSE, warning = FALSE)

library(sp)
library(rgdal)
library(raster)
library(DT)

dir_git <- '~/github/ohi-northeast'
dir_rgn <- file.path(dir_git, 'prep/regions')  ### github: general buffer region shapefiles
dir_anx <- file.path(dir_M, 'git-annex/neprep')

source(file.path(dir_git, '~/github/ohi-northeast/src/R/common.R'))  ### an OHINE specific version of common.R

### goal specific folders and info
goal      <- 'tr'
scenario  <- 'v2016'
dir_goal  <- file.path(dir_git, 'prep', goal, scenario)
dir_goal_anx <- file.path(dir_anx, goal, scenario)

```

#Summary
[general description: What data are being generated? Why (what project, etc.)? Upstream/downstream processing information that might be helpful?  Other information?]

***

#Data Source [NOTE: can be copied from README.md in rawdata file]
**Reference**: [citation for source data; website, literature, contact information. Version of data (if relevant). Screenshots if a series of menus had to be navigated to obtain the data.]

**Downloaded**: [date downloaded or received]

**Description**:  [e.g., surface aragonite state]

**Native data resolution**: [e.g., 1 degree, 30 m, etc.]   

**Time range**: [e.g., 1880-1899, monthly data provided for each year] 

**Format**:  [e.g. NetCDF]

***
  
#Methods
[R code used to generate the data. Or, alternatively, description of the code/data files used to generate the data.]

***

#Citation information  
[citation information: include if these data will have their own specific citation.]