---
title: 'OHI-Northeast: Region Options for the September 21 workshop'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output:
  html_document:
    code_folding: show
    fig_caption: TRUE
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/ohi-northeast/src/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---

*Three different region options are presented here*: 

(1) Whole region, each state (that wants a score), each offshore EPU  
(2) Whole region, state waters north of the Cape, state waters south of the Cape, each offshore EPU  
(3) Whole region, each offshore EPU  


# Load data
```{r setup, include=FALSE, message=F,warning=F}
knitr::opts_chunk$set(fig.width = 10, fig.height = 8, fig.path = 'figs/',
                      echo = FALSE, message = FALSE, warning = FALSE, root.dir = "ohi-northeast")

source('~/github/ohi-northeast/src/R/common.R') ### an OHI-NE specific version of common.R

dir_git <- '~/github/ohi-northeast'
dir_anx <- file.path(dir_M, 'git-annex/neprep')

library(sf)
library(ggplot2)
library(rgeos)
```


## State boundaries and water
```{r}
states <- st_read(dsn = '.',layer = 'states',quiet=T)%>%
          st_transform(p4s_nad83)

coast_st <- states %>%
            filter(!NAME %in% c("Pennsylvania","New Jersey"))

ggplot() +
  geom_sf(data = coast_st,aes(fill = NAME))+
  theme_bw()+ 
  labs(fill='States') +
scale_fill_manual(values = blues9)+
  theme(legend.text=element_text(size=12),
        legend.title = element_text(size = 14))+ 
  coord_sf(crs = st_crs(states), datum = NA) 


states <- st_intersection(states, st_set_crs(st_as_sf(as(raster::extent(-75,-65,39.5,45), "SpatialPolygons")), st_crs(states)))

#state waters with some Maine rivers removed (done in QGIS)
state_wa <- st_read(dsn = file.path(dir_anx, 'spatial'), layer = 'StateWaters_wo_rivs_cleaned',quiet=T)%>%
              st_transform(p4s_nad83)

all_state_wa <- state_wa %>%
                group_by(REGION10) %>%
                summarize()

```

## Northeast Ocean Plan Study Area

Emily emailed this shapefile to us to use as the NE Ocean Plan study area. State waters are not included in this shapefile but we need to include state waters within the larger NE study area.

```{r ne_roi}
#NE ocean plan region of interest - provided by Emily

# ne_roi <- st_read(dsn = file.path(dir_anx,'spatial'),layer = 'ne_plan_poly', quiet = T)%>%
#             st_transform(p4s_nad83)%>%
#           mutate(region = "NE Ocean Plan Study Area") %>%
#          st_cast("MULTIPOLYGON")

#ne study area plus state waters. I had to use this neoceanplanningboundary_polygon shapefile due to weird slivers when mergine the ne_plan_poly with state_wa

# ne <- st_read(dsn = file.path(dir_anx, 'spatial'), layer = 'neoceanplanningboundary_polygon',quiet=T) %>%
#         st_transform(crs = p4s_nad83) %>%
#         group_by(Id) %>%
#         summarize() %>%
#         st_intersection(., eez)
# 
## not sure why this isn't creating just one polygon, so doing a group_by summarize
# ne_sa <- st_union(ne, state_wa) %>%
#           mutate(region = "NE Ocean Plan Study Area")
# 
# st_write(ne_sa, "ne_region_plus_states.shp", driver = "ESRI Shapefile")

ne_sa <- st_read(dsn = ".", layer = "ne_region_plus_states", quiet = T)
```


## Canada EEZ
```{r ca_eez}

# ca_eez  <- st_read(dsn = file.path(dir_M,'git-annex/globalprep/spatial/d2014/data'),layer = 'regions_gcs',quiet=T)%>%
#                     filter(rgn_nam == 'Canada',
#                            rgn_typ == "eez")%>%
#                     st_transform(p4s_nad83)
# 
# ca_eez_crop = st_intersection(ca_eez, st_set_crs(st_as_sf(as(raster::extent(wgs_ext), "SpatialPolygons")), st_crs(ca_eez)))
# 
# st_write(ca_eez_crop, "ca_eez_crop.shp", driver = "ESRI Shapefile")

ca_eez_crop <- st_read('.', "ca_eez_crop", quiet=T)
```

## US EEZ

Also need to grab the tip of maine and add it to the Scotian Shelf EPU
```{r}

eez <- st_read(dsn = '.',layer = 'ne_eez',quiet=T)%>%
        st_transform(crs = p4s_nad83)

maine <- st_intersection(eez, st_set_crs(st_as_sf(as(raster::extent(-68,-66,44,45), "SpatialPolygons")), st_crs(eez)))

```

## Ecological Production Units

The EPU's don't encompass all of Maine's waters up to the border with Canada. I'm using the EEZ to extend the Northernmost EPU to Canada.

```{r epu_extend}

 epu <- st_read(file.path(dir_anx, 'spatial/data_for_rgn_options/Extended_EPU'),'EPU_extended',quiet=T)%>%
         st_transform(p4s_nad83)%>%
         mutate(longname = c('Georges Bank','Gulf of Maine','Scotian Shelf','Mid-Atlantic Bight'))

 #cropping epu extent for plotting
 epu<- st_intersection(epu, st_set_crs(st_as_sf(as(raster::extent(-75,-65,39.5,45), "SpatialPolygons")), st_crs(epu)))

 ss <- epu%>%
         filter(longname == "Scotian Shelf")%>%
         st_union(maine)%>%
         dplyr::select(EPU,Shape_Leng,Shape_Area,longname,geometry)%>%
         st_cast("MULTIPOLYGON")

#join back to EPU

 epu_ext <- epu%>%
            filter(EPU != "SS")%>%
             rbind(ss)%>%
             st_cast("MULTIPOLYGON")
 
 #st_write(epu_ext, "epu_extended.shp", driver = "ESRI Shapefile")

## mask out state waters
s = as(state_wa, 'Spatial')
epu_ext_shp <- as(epu_ext,'Spatial')
 
epu_ext_offshore = epu_ext_shp - s

# mask with states as well
e = epu_ext_offshore - as(states,'Spatial')

#writeOGR(epu_ext_offshore, dsn = ".", layer = "epu_extended", driver = "ESRI Shapefile")

epu_ext_offshore <- st_read('.','epu_extended_offshore',quiet=T)
```

# Option 1

Just the Northeast planning region

```{r option_1}
ggplot() +
  geom_sf(data = states,colour = "lightgray",fill = "beige")+
  geom_sf(data = ne_sa,aes(fill = region))+
  theme_bw()+
  ggtitle("Option 1")+ 
  labs(fill='Regions') +
 scale_fill_manual(values = c("#E0F3F8"))+ 
  theme(legend.text=element_text(size=12),
        legend.title = element_text(size = 14))+ 
  coord_sf(crs = st_crs(states), datum = NA) 


```


# Option 2

Whole region, each offshore EPU 

```{r}
#intersect state waters with EPU to get the massachussets split

all_states <- state_wa%>%
              group_by(REGION10)%>%
              summarize()%>%
              mutate(region = "All State Waters")

```

```{r option_2}
ggplot() +
  geom_sf(data = states,colour = "lightgray",fill = "beige")+
  geom_sf(data = epu_ext,aes(fill = longname))+
  geom_sf(data = ne_sa,aes(fill = region),alpha = 0, lwd = 0.5)+
  theme_bw()+
  ggtitle("Option 2")+ 
  labs(fill='Regions') +
 scale_fill_manual(values = c("#313695", "#ABD9E9", "#E0F3F8","#FFFFFF","#4575B4"))+ 
  theme(legend.text=element_text(size=12),
        legend.title = element_text(size = 14))+ 
  coord_sf(crs = st_crs(states), datum = NA) 
```


# Option 3

Whole region, each state (that wants a score), each offshore EPU  

```{r option_3}
ggplot() +
  geom_sf(data = states,colour = "lightgray",fill = "beige")+
  geom_sf(data = ne_sa,aes(fill = region))+
  geom_sf(data = state_wa,aes(fill = NAME10))+
  theme_bw()+
  ggtitle("Option 3")+ 
  labs(fill='Regions') +
 scale_fill_manual(values = c("#A50026",  "#ABD9E9","#313695", "#E0F3F8","#D73027","#FEE090" ,"#F46D43",  "#FDAE61", "#74ADD1", "#4575B4"))+ 
  theme(legend.text=element_text(size=12),
        legend.title = element_text(size = 14))+ 
  coord_sf(crs = st_crs(states), datum = NA) 
```

# Option 4

Whole region, state waters north of the Cape, state waters south of the Cape, each offshore EPU  

```{r}
#intersect state waters with EPU to get the massachussets split

int <- st_intersection(epu_ext,state_wa)

north <- int%>%
          filter(NAME10 %in% c("Maine","New Hampshire","Massachusetts"),
                 EPU %in% c("GOM", "SS"))%>%
          mutate(region = "State waters N of Cape Cod")%>%
          group_by(region)%>%
          summarize()

#st_write(north, "n_cape_cod.shp", driver = "ESRI Shapefile")

south <- int%>%
          filter(NAME10 %in% c("New York","Rhode Island","Connecticut","Massachusetts"),
                 EPU %in% c("MAB","GB"))%>%
          mutate(region = "State waters S of Cape Cod")%>%
          group_by(region)%>%
          summarize()

#st_write(south, "s_cape_cod.shp", driver = "ESRI Shapefile")

```

```{r option_4}

ggplot() +
  geom_sf(data = states,colour = "lightgray",fill = "beige")+
  geom_sf(data = epu_ext,aes(fill = longname))+
  geom_sf(data = north,aes(fill = region))+
  geom_sf(data = south,aes(fill = region))+
  geom_sf(data = ne_sa,aes(fill = region),alpha = 0, lwd = 0.5)+
  theme_bw()+
  ggtitle("Option 4")+ 
  labs(fill='Regions') +
 scale_fill_manual(values = c("#313695", "#ABD9E9", "#E0F3F8","#FFFFFF","#4575B4","#FEE090","#FDAE61"))+ 
  theme(legend.text=element_text(size=12),
        legend.title = element_text(size = 14))+ 
  coord_sf(crs = st_crs(states), datum = NA) 
```


# Overlap of Canada

```{r canada_overlap}

#cropping epu extent for plotting
canada <- st_intersection(ca_eez_crop, st_set_crs(st_as_sf(as(raster::extent(-75,-65,39.5,45), "SpatialPolygons")), st_crs(ca_eez_crop)))

ggplot() +
  geom_sf(data = states,colour = "lightgray",fill = "beige")+
  geom_sf(data = epu_ext,aes(fill = longname))+
  geom_sf(data = canada,aes(fill = rgn_nam,alpha = 0.4),show.legend = F)+
  geom_sf(data = all_states,aes(fill = region))+
  theme_bw()+
  ggtitle("Overlapping Canada EEZ with EPU")+ 
  labs(fill='Regions') +
 scale_fill_manual(values = c("#FEE090","coral2","#313695", "#ABD9E9", "#E0F3F8","#4575B4"))+ 
  theme(legend.text=element_text(size=12),
        legend.title = element_text(size = 14))+ 
  coord_sf(crs = st_crs(states), datum = NA) 

```

# Aggregate EPUS



```{r agg_epus}

epu_combined <- st_union(epu_ext,by_feature = F)#union all epu units together

df <- data.frame(region = "Offshore") #create dataframe to join to epu_combined (it's of class `sfc` so no dataframe yet)

#I was struggling with how to do this so I copied the example in ?st_geometry. 
st_geometry(df)<-epu_combined 
class(df)
st_geometry(df)
st_geometry(df) <- epu_combined
st_geometry(df) <- NULL
#it works!
sf <- st_set_geometry(df,epu_combined)

                

ggplot() +
  geom_sf(data = states,colour = "lightgray",fill = "beige")+
  geom_sf(data = sf,aes(fill = region))+
  geom_sf(data = state_wa,aes(fill = NAME10))+
  theme_bw()+
  ggtitle("Option 4")+ 
  labs(fill='Regions') +
 scale_fill_manual(values = c("#ABD9E9", "#FEE090"  ,"#313695","#A50026","#D73027","#E0F3F8" ,"#F46D43",  "#FDAE61", "#74ADD1", "#4575B4"))+ 
  theme(legend.text=element_text(size=12),
        legend.title = element_text(size = 14))+ 
  coord_sf(crs = st_crs(states), datum = NA) 

```

```{r}

ggplot() +
  geom_sf(data = states,colour = "lightgray",fill = "beige")+
  geom_sf(data = sf,aes(fill = region))+
  geom_sf(data = all_states,aes(fill = region))+
  theme_bw()+
  ggtitle("Option 5")+ 
  labs(fill='Regions') +
 scale_fill_manual(values = c("#FEE090","#ABD9E9"))+ 
  theme(legend.text=element_text(size=12),
        legend.title = element_text(size = 14))+ 
  coord_sf(crs = st_crs(states), datum = NA) 

```


# Coastal County and State Map

```{r coastal_counties}

counties <- st_read(dsn = '.',layer = 'ne_counties',quiet=T)%>%
          st_transform(p4s_nad83)

#reading in states again. Yes this is redundant but it includes NJ and PA and is prettier for mapping..
states <- st_read(dsn = '.',layer = 'states',quiet=T)%>%
          st_transform(p4s_nad83)


ggplot() +
  geom_sf(data = states,colour = "lightgray",fill = "beige")+
  geom_sf(data = counties,aes(fill = NAME_1))+
  theme_bw()+ 
  labs(fill='Counties') +
scale_fill_manual(values = blues9)+
  theme(legend.text=element_text(size=12),
        legend.title = element_text(size = 14))+ 
  coord_sf(crs = st_crs(states), datum = NA) 
```

