---
title: 'OHI-Northeast: Tourism and Recreation'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/ohi-northeast/src/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---

#Setup
``` {r setup, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'Figs/',
                      echo = FALSE, message = FALSE, warning = FALSE)

library(zoo)
library(rgdal)
library(raster)
library(DT)

dir_git <- '~/github/ohi-northeast'
dir_rgn <- file.path(dir_git, 'prep/regions')  ### github: general buffer region shapefiles
dir_anx <- file.path(dir_M, 'git-annex/neprep')

source(file.path(dir_git, 'src/R/common.R'))  ### an OHIBC specific version of common.R


### set up proj4string options: BC Albers and WGS84
p4s_wgs84 <- '+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0'
p4s_nealb <- '+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=23 +lon_0=-96 +x_0=0 +y_0=0 +ellps=clrk66 +units=m +no_defs'

```

#Summary
[general description: What data are being generated? Why (what project, etc.)? Upstream/downstream processing information that might be helpful?  Other information?]

***



#Data Source [NOTE: can be copied from README.md in rawdata file]
**Reference**: [citation for source data; website, literature, contact information. Version of data (if relevant). Screenshots if a series of menus had to be navigated to obtain the data.]

**Downloaded**: [date downloaded or received]

**Description**:  [e.g., surface aragonite state]

**Native data resolution**: [e.g., 1 degree, 30 m, etc.]   

**Time range**: [e.g., 1880-1899, monthly data provided for each year] 

**Format**:  [e.g. NetCDF]

***
  
  
  
#Methods

```{r read in data}

#read in raw data and filter for Sector == Tourism & Recreation
raw <- read.csv(file.path(dir_anx,"_raw_data/NOEP/NewEngland_sectors_counties.csv"),stringsAsFactors=F)%>%
        filter(Sector=='Tourism & Recreation')%>%
      arrange(State,County)%>%
        mutate(year = rep(2013:2005,27),
               Employment = as.numeric(gsub(",","",.$Employment)),
               Establishments = as.numeric(gsub(",","",.$Establishments)))

```

Calculate the number of jobs over time

```{r jobs}

jobs <- raw%>%
          group_by(State, year)%>%
          summarise(jobs = sum(Employment,na.rm=T))%>%
          ungroup()%>%
          as.data.frame()%>%
          mutate(roll_avg = rollapply(jobs, 5,FUN=mean,align = 'right',partial=T),
                 score = jobs/roll_avg)

```


***

#Citation information  

National Ocean Economics Program. Ocean Economic Data by Sector & Industry., ONLINE. 2012.
Available: http://www.OceanEconomics.org/Market/oceanEcon.asp [29 Apr. 2016]

``` {r child = file.path(dir_git, 'src/templates/ohibc_prov_ftr.Rmd')}

