---
title: 'OHI-Northeast: Tourism and Recreation'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/ohi-northeast/src/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---

``` {r setup, echo=F, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(fig.width = 8, fig.height = 6, fig.path = 'figs/', message = FALSE, warning = FALSE, echo=F)
dir_git <- '~/github/ohi-northeast'
dir_rgn <- file.path(dir_git, 'prep/regions')  ### github: general buffer region shapefiles
source(file.path(dir_git, 'src/R/common.R'))  ### an OHIBC specific version of common.R

library(zoo)
library(rgdal)
library(raster)
library(DT)
library(plotly)

dir_anx <- file.path(dir_M, 'git-annex/neprep')

#round to 2 decimals
options(digits=2)

```

#Summary

The Tourism & Recreation goal is calculated using coastal employment data from the [National Ocean Economics Program](http://www.oceaneconomics.org/), and State-wide employment from the [Bureau of Labor Statistics](http://www.bls.gov/data/).

***

#Data Sources

##National ocean Economics Program (NOEP)

**Downloaded**: Received from Pat Johston on May 17, 2016.  
**Description**:  Total number of jobs per sector for RI, ME, MA, CT and NH counties from 2005 to 2013. The data also include number of establishments, GDP and Wages for each sector - state - year. 
**Native data resolution**: County level   
**Time range**: 2005 - 2013  
**Format**:  Tabular  


```{r coastal_data}

#read in raw data and filter for Sector == Tourism & Recreation
data<- read.csv(file.path(dir_anx,"_raw_data/NOEP/NewEngland_sectors_counties-2.csv"),
                stringsAsFactors=F)%>%
        filter(Sector=='Tourism & Recreation')%>%
        mutate(Employment = as.numeric(gsub(",","",.$Employment)),
               Establishments = as.numeric(gsub(",","",.$Establishments)))

DT::datatable(data, rownames=F, caption = 'Employment in the Tourism & Recreation sector of Coastal Jobs in New England provided by NOEP')

```



##US Bureau of Labor Statistics

**Downloaded**: May 17, 2016  
**Description**: Total Annual Employment (in thousands) per State from 1990 to 2015  
**Native data resolution**: State level  
**Time Range:** 2005 to 2015  
**Format:** Tabular  


```{r state data}
#read in state-wide employment. This data was made by hand from a manual query on the BLS.gov site: http://data.bls.gov/pdq/SurveyOutputServlet. 
# this should be replaced with code using the blsAPI R package. Currently waiting on a response from the maintainer: https://github.com/mikeasilva/blsAPI/issues/7

#employment is in thousands so we need to multiply
st_emp <- read.csv('state_employment.csv')%>%
          mutate(Employment = 1000*Employment)

DT::datatable(st_emp, caption = 'Total Employment by State, provided by the Bureau of Labor Statistics',rownames=FALSE)


```

***

#Methods

The Tourism and Recreation score, $X_{i}$, is calculated using the following equation; 

$$X_{i} = \frac{\frac{C_{i}}{C_{r}}}{\frac{T_{i}}{T_{r}}}$$

where $C_{i}$ is the number of coastal jobs in year *i*, $C_{r}$ is the number of coastal jobs in 2005, ${T_{i}}$ is the total number of jobs statewide for year *i* and ${T_{r}}$ is the number of jobs statewide in 2005.

```{r jobs}

jobs<- data%>%
          filter(!grepl("All",County))%>%
          group_by(State, Year)%>%
          summarise(coast_jobs = sum(Employment,na.rm=T))%>%
          mutate(hist_jobs = coast_jobs[which.min(Year)])%>%
          left_join(st_emp,by=c('State','Year'))%>%
          mutate(cst_chg = coast_jobs/hist_jobs, #use historical reference point from 2005
                 hist_st_jobs = Employment[which.min(Year)],
                 st_chg = Employment/hist_st_jobs,
                 score = ifelse((cst_chg/st_chg)>1,1,cst_chg/st_chg))%>%as.data.frame()

```

***

#Results

```{r table}

jobs = filter(jobs,!is.na(score))

write.csv(jobs, 'tr_scores.csv')

DT::datatable(jobs,colnames = c('State','Year','Coastal Jobs','State Jobs','Coastal 5 yr mean','Coastal job change','State 5 yr mean','State job change','Final Score'), caption = 'Coastal employment in the Tourism & Recreation sector for all five regions as well as total number of jobs per region. The 5 year means are calculated using the previous 5 years data and the change in number of jobs is calculated using current jobs divided by the 5 year mean.', rownames=F)

```

```{r plot1}

tr_plot <- ggplot(jobs,aes(x = Year,y = score, group=State))+
  ggtheme_plot+
  geom_point(aes(color = State)) +
            geom_line(aes(color = State))+
            scale_y_continuous(limits = c(0.75,1))+
            ggtitle("Tourism & Recreation Scores for the Northeast Regions")+
  labs(y = 'Score')

ggplotly(tr_plot)


ggsave('tr_scores.png')            
```


***

#Citation information  

National Ocean Economics Program. Ocean Economic Data by Sector & Industry., ONLINE. 2012.
Available: http://www.OceanEconomics.org/Market/oceanEcon.asp [17 May 2016]

Bureau of Labor Statistics, U.S. Department of Labor, State and Metro Area Employment, Hours & Earnings. 5/17/2016. [www.bls.gov/SAE/](http://www.bls.gov/sae/).
