---
title: 'OHI-Northeast: Livelihoods'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/ohi-northeast/src/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---


``` {r setup, echo=F, message = FALSE, warning = FALSE, results = 'hide'}
knitr::opts_chunk$set(fig.width = 10, fig.height = 6, fig.path = 'figs/', message = FALSE, warning = FALSE)

dir_git <- '~/github/ohi-northeast'
source(file.path(dir_git, 'src/R/common.R'))  ### an OHI-NE specific version of common.R

library(tidyverse)
library(readxl)
library(DT)
#install.packages("striprtf")
library(striprtf)
library(zoo)

dir_anx <- file.path(dir_M, 'git-annex/neprep')

```

#Summary

The Livelihoods sub-goal is calculated using Ocean Economy employment and wage data from the [National Ocean Economics Program](http://www.oceaneconomics.org/). There are two components to this goal model, **jobs (or employment)** and **wages**. NOEP provides data on the number of jobs, and total wages for jobs that directly or indirectly depend upon the ocean. Annual job and wage growth is measured by comparing total employment and mean annual wages to the average employment and wages of the previous year. The target for region job growth is set to be equal to or greater than the national average, calculated using data from the Bureau of Labor Statistics. The target for mean annual wage growth is set at 3.5%, reflecting the [*Nominal Wage Growth Target*](http://www.epi.org/nominal-wage-tracker/) as set by the Federal Reserve.

***

# Data Source

## National ocean Economics Program (NOEP)

**Downloaded**: Received from Pat Johston on October 3, 2017.  
**Description**:  Total number of jobs and wages per sector for RI, ME, MA, CT, NY and NH counties from 2005 to 2013. The data also include number of establishments and GDP for each sector - state - year.  
**Native data resolution**: County level     
**Time range**: 2005 - 2014  
**Format**:  Tabular  

***

# Jobs

The livelihoods jobs score, $j '$, is calculated using the following equation; 

$$j ' = \frac{\frac{C_{i}}{C_{r}}}{\frac{A_{i}}{A_{i-1}}}$$

where $C_{i}$ is the number of coastal jobs in year *i*, $C_{i-1}$ is the average number of coastal jobs for the previous year, ${A_{i}}$ is the total number of jobs nationwide for year *i* and ${A_{i-1}}$ is the average number of jobs nationwide for the previous year.


## Coastal jobs

Two sets of data from NOEP are provided, one for the timeframe 1990 - 2004 and the other from 2005 - 2014. NOEP recommends not using these two together due to inconsistincies across the two, especially in regards to GDP. While both are read in here, we only use the 2005 - 2014 data to calculate the LIV goal for now.

```{r coastal_data}

## Read in raw data which came in .xlsx format with 2 sheets. 

## read in NOEP sheet for all data 1990 - 2004
noep_data_1 <- read_excel(file.path(dir_anx, '_raw_data/NOEP/New_England_ocean_series.xlsx'), sheet = "NOEP")
  
## read in ENOW sheet which contains data for 2005 - 2014 
noep_data_2 <- read_excel(file.path(dir_anx, '_raw_data/NOEP/New_England_ocean_series.xlsx'), sheet = "ENOW")
```

Moving forward just with the more recent data series. Data is provided for each individual coastal county as well as "All Counties" in each state. There is an important distinction between using "All Counties" and simply summing up all state county data. Some data at the county level is supressed due to regulation. But when reported at the State level, this data may be released. We see this when looking at New Hampshires marine transportation data. The sum of NH's two coastal counties for jobs does not equal the reported value for jobs for "All New Hampshire Counties". This is most likely due to data supression at the county level. Therefore we use the sum of all counties rather than the data reported for "All Counties".

```{r split_MA}
# need to read in the MA counties since they are divided across two bioregions

ma_ctys <- read_csv("~/github/ohi-northeast/src/tables/MA_counties.csv")

clean_data <- noep_data_2 %>%
       filter(Sector == 'All Ocean Sectors') %>%
        left_join(rgn_data, by = c("State" = "rgn_name")) %>%
        left_join(ma_ctys, by = c("State", "County")) %>%
        mutate(rgn_id = ifelse(is.na(rgn_id.x), rgn_id.y, rgn_id.x),
               rgn_name = ifelse(is.na(rgn_name), State, rgn_name)) %>%
  select(-rgn_id.x, -rgn_id.y, -area_km2.x, -area_km2.y, -rgn_abrev.x, -rgn_abrev.y) %>%
        mutate(Employment     = ifelse(rgn_name %in% c("Massachusetts-Gulf of Maine", "Massachusetts-Virginian"), 
                                      as.numeric(gsub(",","",.$Employment))/2, as.numeric(gsub(",","",.$Employment))),
               Establishments = ifelse(rgn_name %in% c("Massachusetts-Gulf of Maine", "Massachusetts-Virginian"), 
                                       as.numeric(gsub(",","",.$Establishments))/2, as.numeric(gsub(",","",.$Establishments))),
               Wages          = ifelse(rgn_name %in% c("Massachusetts-Gulf of Maine", "Massachusetts-Virginian"), 
                                       as.numeric(gsub(",","",.$Wages))/2, as.numeric(gsub(",","",.$Wages))),
               Wages_2012     = ifelse(rgn_name %in% c("Massachusetts-Gulf of Maine", "Massachusetts-Virginian"),
                                       as.numeric(gsub(",","",.$Wages_2012))/2, as.numeric(gsub(",","",.$Wages_2012))),
               GDP            = ifelse(rgn_name %in% c("Massachusetts-Gulf of Maine", "Massachusetts-Virginian"),
                                       as.numeric(gsub(",","",.$GDP))/2, as.numeric(gsub(",","",.$GDP))),
               GDP_2012       = ifelse(rgn_name %in% c("Massachusetts-Gulf of Maine", "Massachusetts-Virginian"),
                                       as.numeric(gsub(",","",.$GDP_2012))/2, as.numeric(gsub(",","",.$GDP_2012))))

DT::datatable(clean_data, rownames=F, caption = 'Employment in the Tourism & Recreation sector of Coastal Jobs in New England provided by NOEP')

```

### Meta-analysis

To identify inconsistencies in the data, we take a look at the reported employment values at both the county level and statewide. One would expect that the sum of the county employment values would equal what is reported for the "Statewide" employment values. It seems that this is not the case.

```{r noep_state_vs_county_totals}

states <-  c("Maine", "New Hampshire", "Rhode Island", "Massachusetts", "New York", "Connecticut")

meta <- function(state){
  
  all <- clean_data %>%
    filter(State == !!state,
           str_detect(County, "All")) %>%
    select(Year, Employment) %>%
    rename(all_ctys_employment = Employment)
  
  out <- clean_data %>%
     filter(State == !!state,
            str_detect(County, "All") == FALSE) %>%
     select(State, County, Year, Employment) %>%
     group_by(Year) %>%
    summarize(totals = sum(Employment, na.rm = T)) %>%
    left_join(all) %>%
    rename(county_totals = totals,
           statewide = all_ctys_employment) %>%
    gather(key = spatial_res, value = Employment, -Year) %>%
    mutate(State = state)
  
return(out)
}

t <- map_df(states, meta) %>%
  distinct()

ggplot(t, aes(x = Year, y = Employment, color = spatial_res))+
    geom_line() +
    facet_wrap(~State, scales = "free") +
  scale_color_manual(" ", labels = c("Sum of county reported employment", "Statewide reported employment"), values = c("blue", "red"))
```

There are some clear discrepancies in the dataset between the total number of jobs reported at the state level ("statewide") and the sum of all employment numbers at the County level. New york and Massachusetts have parallel trends, so using either data stream should result in the same final score. New Hampshire has significantly more statewide jobs beginning in 2011 due to data supression up until that point. We can assume that if data were not suppressed pre-2011 statewide, we would also see parallel trends in New Hampshire as well. Operating under that assumption, we can use the county level data. Rhode Island, Connecticut and Maine show low employment numbers in earlier years when adding up at the county level. This could be due to a lack of data. For example, Saghadoc county in Maine has no data up until 2010, when the jump happens. This suggests we should use the statewide data for Maine. There is no missing data in Connecticut or Rhode Island. This might suggest data suppression in those earlier years at the county level. Therefore the statewide data should be used. 


```{r combine_noep_state_county_data}

#select the data for ME, CT and RI, which is going to use the data reported for "All x counties"
state_data <- clean_data %>%
  filter(str_detect(County, "All"),
         State %in% c("Maine", "Connecticut", "Rhode Island")) %>%
  select(state = State, year = Year, rgn_id, rgn_name, rgn_employment = Employment, rgn_wages = Wages_2012, rgn_estab = Establishments)
  
county_data <- clean_data %>%
  filter(str_detect(County, "All")== FALSE,
         State %in% c("Massachusetts", "New York", "New Hampshire")) %>%
  group_by(rgn_id, Year) %>%
  mutate(rgn_employment = sum(Employment, na.rm = T),
         rgn_wages = sum(Wages_2012, na.rm = T),
         rgn_estab = sum(Establishments, na.rm = T)) %>% #employment by region
  select(state = State, year = Year, rgn_id, rgn_name, rgn_employment, rgn_wages, rgn_estab) %>%
  distinct()

coast_jobs <- bind_rows(state_data, county_data) %>%
  select(-rgn_wages, -rgn_estab)

write.csv(coast_jobs, file.path(dir_calc, "layers/le_coast_jobs.csv"))
```


```{r rgn_coastal_employment}

ggplot(coast_jobs, aes(x = year, y = rgn_employment)) +
  geom_line()+
  facet_wrap(~ rgn_name, scales = "free") +
  ylab("Number of jobs") + 
  xlab("Year") +
  ggtitle("Total employment in coastal jobs")

```

<br>

***

<br>


## Reference point

Our reference point is the average number of jobs over the previous year.

```{r jobs_ref_point}

jobs_cst_ref <- coast_jobs %>%
  arrange(year) %>%
  group_by(rgn_id) %>%
  mutate(coast_jobs_1yr = rollapply(rgn_employment, 1, FUN=mean, align = "right", na.rm=F, partial = T),
         coast_jobs_prev_1yr = lag(coast_jobs_1yr, n = 1)) %>%
  select(state, year, rgn_id, rgn_name, ref = coast_jobs_prev_1yr)

write.csv(jobs_cst_ref, file.path(dir_calc, "layers/le_coast_jobs_ref.csv"))

```


**Jobs directly related to ocean in coastal counties**

```{r coastal_jobs_and_ref_point}
c <- coast_jobs %>%
  left_join(jobs_cst_ref) %>%
  select(rgn_id, year, rgn_employment, ref, rgn_name) %>%
  gather(data, jobs, -rgn_id,-year, -rgn_name)

ggplot(c, aes(x = year, y = jobs, color = data)) +
  geom_line() +
  facet_wrap(~rgn_name, scales = "free") +
  ylab("Number of jobs") +
  xlab("Year") +
  ggtitle("Regional employment in coastal jobs") +
  scale_color_manual(" ", labels = c("Mean employment over the \nprevious 3 years","Annual employment"), values = c("red", "blue"))
```

***

# Wages

The livelihoods wages score, $g'$ is calculated using the following equation;

$$g ' = \frac{\frac{W_{i}}{W_{i-1}}}{{1.035}}$$

where $W_{i}$ is the mean annual wage for regional coastal employment in year *i*, and $W_{i-1}$ is the mean wage for coastal employment for the previous year. The change in wages (${\frac{W_{i}}{W_{i-1}}}$) is then divided by the target, 1.035, representing the 3.5% Nominal Wage Growth Target established by the Federal Reserve.

## Coastal wages

We already cleaned up this NOEP dataset above and can just create a new dataframe, `coast_wages` by selecting the `rgn_wages` column rather than `rgn_employment`. Wages are also standardized by total employment to give us a measure of wages per capita as opposed to total wages.

```{r coast_wages}

coast_wages <- bind_rows(state_data, county_data) %>%
  mutate(wages_per_job = rgn_wages/rgn_employment) %>%
  select(state, year, rgn_id, rgn_name, wages_per_job)

write.csv(coast_wages, file.path(dir_calc, "layers/le_coast_wages.csv"))

```


```{r coastal_wages_plot}
ggplot(coast_wages, aes(x = year, y = wages_per_job)) +
  geom_line() +
  ylab("Wages (2012 USD$)") +
  facet_wrap(~ rgn_name, scales = "free") +
  ggtitle("Wages per capita") +
  xlab("Year")
```

## Reference point

Our reference point is the mean annual wage per job over the previous 3 years.

```{r coastal_wages_ref_points}

wages_cst_ref <- coast_wages %>%
  arrange(year) %>%
  group_by(rgn_id) %>%
  mutate(coast_wages_1yr = rollapply(wages_per_job, 1, FUN=mean, align = "right", na.rm=F, partial = T),
         coast_wages_prev_1yr = lag(coast_wages_1yr, n = 1)) %>%
  select(state, year, rgn_id, rgn_name, ref = coast_wages_prev_1yr)

write.csv(wages_cst_ref, file.path(dir_calc, "layers/le_coast_wages_ref.csv"))
```

**Wages per job directly related to ocean in coastal counties**

```{r coastal_wages_and_ref_point}

c <- coast_wages %>%
  left_join(wages_cst_ref) %>%
  select(rgn_id, year, wages_per_job, ref, rgn_name) %>%
  gather(data, wages, -rgn_id,-year, -rgn_name)

ggplot(c, aes(x = year, y = wages, color = data)) +
  geom_line() +
  facet_wrap(~rgn_name, scales = "free") +
  ylab("Wages per job (2012 USD$") +
  xlab("Year") +
  ggtitle("Average annual wages for coastal jobs") +
  scale_color_manual(" ", labels = c("Mean wages for the \nprevious year","Annual average wages"), values = c("red", "blue"))
```


***

#Setting targets

Using Bureau of Labor Statistics data, we can get the nationwide average job growth and use this as our target for coastal job growth.

```{r}
#devtools::install_github("keberwein/blscrapeR")
library(blscrapeR)
#read in API access key that is saved on Mazu
blsKey <- read_rtf(file.path(dir_M,'git-annex/neprep/keys/BureauofLaborStatistics.rtf'))
set_bls_key(blsKey, overwrite=TRUE)
```

```{r bls_us_jobs_stats}

us_employment <- bls_api("ENUUS00010010",
              startyear = 2001, endyear = 2016, registrationKey = "BLS_KEY", annualaverage=TRUE) %>%
  as.data.frame() %>%
  filter(periodName == "Annual") %>%
  select(year, us_jobs = value) %>%
  arrange(year) %>%
  mutate(us_jobs_1yr = rollapply(us_jobs, 1, FUN=mean, align = "right", na.rm=F, partial = T),
         us_jobs_prev_1yr = lag(us_jobs_1yr, n = 1),
         us_job_growth = us_jobs/us_jobs_prev_1yr) %>%
  select(year, us_job_growth)

ggplot(us_employment, aes(x = year, y = us_job_growth)) +
  geom_line() +
  ylab("Job growth compared to previous year")

```

#Results

While this is done in the OHI toolbox, the calculated results are duplicated here.

## Jobs

```{r jobs_score}
  ##combining the employment data and calculating the change compared to reference period
  
  ##coastal jobs
  cst_jobs <- coast_jobs %>%
    rename(coast_jobs = rgn_employment) %>%
    left_join(jobs_cst_ref) %>%    #join with the reference point data
    filter(!is.na(ref)) %>%     #remove years with NA for mean jobs (same as removing years pre-2010)
    rename(coast_mean_jobs = ref) %>%
    mutate(cst_chg = coast_jobs/coast_mean_jobs)

  #combine coastal and state data, calculate jobs scores
  jobs_score <- cst_jobs %>%
    left_join(us_employment) %>%        #add in the national statistics for job growth
    mutate(job_score = ifelse(cst_chg >= us_job_growth, 1, cst_chg/us_job_growth)*100) %>% 
    select(rgn_id, year, job_score) %>%
    left_join(rgn_data)
  
  ggplot(jobs_score, aes(x = year, y = job_score, color = rgn_name)) +
    geom_line() +
      ylab("Score")+
      ggtitle("Jobs score")
```


## Wages 

According to the Federal Reserve, the *Nominal Wage Growth Target* is 3 - 3.5% annually. [EPI report explaining this in more detail](http://www.epi.org/nominal-wage-tracker/).

```{r wages_scores}
  
  ## Wages scores
  #combining the coastal wage data and calculate the change
  cst_wages <- coast_wages %>%
    left_join(wages_cst_ref) %>%    #join with the reference point data
    filter(!is.na(ref)) %>%       #remove years with NA for wages (same as removing years pre-2010)
    rename(coast_mean_wages = ref) %>%
    mutate(cst_chg = wages_per_job/coast_mean_wages)

  
  #combine coastal and state data, calculate
  wages_score <- cst_wages %>%
    mutate(wages_score = ifelse(cst_chg > 1.035, 1, cst_chg/1.035)*100) %>%
    select(rgn_id, year, wages_score) %>%
    left_join(rgn_data)
  
    ggplot(wages_score, aes(x = year, y = wages_score, color = rgn_name)) +
    geom_line() +
      ylab("Score")+
      ggtitle("Wages score")
```  

## Livelihoods status

Combine jobs and wages scores per region and divide by two.
```{r liv_status}  
  # status
  liv_status <- jobs_score %>%
    left_join(wages_score) %>%
    mutate(score = (job_score + wages_score)/2,
           dimension = "status") %>%
    select(year, region_id = rgn_id, score, dimension) %>%
  left_join(rgn_data, by = c("region_id" = "rgn_id"))

ggplot(liv_status, aes(x = year, y = score, color = rgn_name))+
  geom_line() +
  ggtitle("Livelihoods subgoal status scores") +
  ylab("Score") +
  xlab("Year")

```


***

#References

National Ocean Economics Program. Ocean Economic Data by Sector & Industry., ONLINE. 2012.
Available: http://www.OceanEconomics.org/Market/oceanEcon.asp [17 May 2016]


Bureau of Labor Statistics, U.S. Department of Labor, Quarterly Census of Employment and Wages. 7/24/2016. http://www.bls.gov/cew/](http://www.bls.gov/cew/).